A Gentle Introduction into Structural Causal Models

### 2021-06-01 draft.Rmd
draft

Most students have heard the phrase 'correlation does not imply causation.'
While correlation implies co-occurrence, for many problems correlation is not enough. 
Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
For many problems we want to understand causal relationships between variables.
There are different approaches on how to model causal relationships.
This paper focuses on Structural Causal Models and Bayesian Causal Networks.
Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
Summarized, *'they are simulators that generate a distribution based on noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
- noise variables help generate joint distribution [@barocas].
SCMs are non-parametric structural equations models with added features.
Existing literature has provided an excellent introduction to SCMs and BCNs. 
@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
His work addressed misconceptions in social sciences, causality and statistics.
His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
Their work highlights the importance of time in causality and inference based on observational data alone. 
E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
@tarka2018overview discuss the history of SEMs and SCMs.
The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
I structure the rest of the paper as follows: 
Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
Section 3 addresses Pearls Causal Hierachy

@@

Disturbance in SCM: Correlated and causal factor ; responsible for variation
Disturbance in Regression: Uncorrelated

@@

exploit invariant characteristics of SEM whithout committing to shape. 
structural if function autonomous and invariant to change in form of other functions.

@@

Another issue in this paper is understanding the relationship of time and causality.
SCMs make the assumption that causal relationships hold over time[@peters_elements_2017].
These vague definitions of time is more prevalent in social sciences. 
Meanwhile hard sciences deal with time in a more concise manner. 
Differential equations model time on a more mechanic manner [@mooij2013ordinary].

$$
\begin{array}{|l|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { predict in IID } \\
\text { setting }
\end{array} & \begin{array}{l}
\text { predict under } \\
\text { changing } \\
\text { distributions / } \\
\text { interventions }
\end{array} & \begin{array}{l}
\text { answer } \\
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { obtain } \\
\text { physical } \\
\text { insight }
\end{array} & \begin{array}{l}
\text { automatically } \\
\text { learn from } \\
\text { data }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model } \\
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & ? \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{Y} ? ? \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{Y} ? \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} & \mathrm{Y} \\
\hline
\end{array}
$$

>>>>

Most students have heard the phrase 'correlation does not imply causation.'
While correlation implies co-occurrence, for many problems correlation is not enough. 
Algorithmic decision making based on co-occurrence is insufficient  in high stake settings [@bareinboim2020pearl].
For many problems we want to understand causal relationships between variables.
There are different approaches on how to model causal relationships.
This paper focuses on Structural Causal Models and Bayesian Causal Networks.
Bayesian Causal Networks (BCN) cast a model based on **conditional probabilities.**
Structural Causal Models (SCM) specify relationships based on **functional equations** [@pearl2009causality].
Summarized, *'they are simulators that generate a distribution based on noise variables (independent) and a unique joint distribution...'* [@hardtrecht].
- noise variables help generate joint distribution [@barocas].

- latent factors great [@creager][@oberst2019counterfactual].
- untestable assumptions [@pearl2009causality] frowned upon by statistics for long time
- graphical framework better for modeling [@morgan_winship_2014] and reason why this appraoch is better than others

SCMs are non-parametric structural equations models with added features.
Existing literature has provided an excellent introduction to SCMs and BCNs. 
@pearl2009causality provided a comprehensive introduction into this topic with his book 'Causality'.
His work addressed misconceptions in social sciences, causality and statistics.
His contribution on the hierarchy of causation advanced the comprehensive understanding of causality. 
@peters_elements_2017 address the relationship of causality, physical sciences, and causal inference.
Their work highlights the importance of time in causality and inference based on observational data alone. 
E.g. [@bareinboim2020pearl] worked on the application of causal methods in machine learning. 
@tarka2018overview discuss the history of SEMs and SCMs.
The aim of this paper is to summarise SCMs and it's intersection with social sciences and physics.
I structure the rest of the paper as follows: 
Section 2 introduces the assumptions in causal modelling. Further, section 2 discusses different causal differencecs, fundamental differences and a brief history of the methods. 
Section 3 addresses Pearls Causal Hierachy

@@

Disturbance in SCM: Correlated and causal factor ; responsible for variation
Disturbance in Regression: Uncorrelated [@pearl2009causality]

@@

exploit invariant characteristics of SEM without committing to shape. 
structural if function autonomous and invariant to change in form of other functions.

Example:

(#TODO change to own example)
Joint factorization sprinker

$$
P_{X_{3}=\mathrm{On}}\left(x_{1}, x_{2}, x_{4}, x_{5}\right)=P\left(x_{1}\right) P\left(x_{2} \mid x_{1}\right) P\left(x_{4} \mid x_{2}, X_{3}=\mathrm{On}\right) P\left(x_{5} \mid x_{4}\right)
$$

$$
\begin{array}{l}
P_{x}(v)=\prod_{\left\{i \mid V_{i} \notin X\right\}} P\left(v_{i} \mid p a_{i}\right)\\
\text { for all } \boldsymbol{v} \text { consistent with } x
\end{array}
$$

@@

Another issue in this paper is understanding the relationship of time and causality.
SCMs make the assumption that causal relationships hold over time[@peters_elements_2017].
These vague definitions of time is more prevalent in social sciences. 
Meanwhile hard sciences deal with time in a more concise manner. 
Differential equations model time on a more mechanic manner [@mooij2013ordinary].
scriptsize

$$
\begin{array}{|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { IID setting }
\end{array} & \begin{array}{l}
\text { changing } \\
\text { distributions } 
\end{array} & \begin{array}{c}
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { physical } \\
\text { insight }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
\hline
\end{array}
$$
